\Large Generative Dynamics of Supreme Court Citations

### 2018-05-02 Tex/paper.tex
revised the covariate discussion

 \section{Empirical Analysis}
Our three data sources for this study include the Supreme Court Database (SCDB) \citep{spaeth2014supreme}, Martin-Quinn scores \citep{martin2002dynamic}, and Supreme Court citation data provided by the CourtListener Free Law Project \citep{CourtListener}. In the next section we explain the variables we construct using these data sources. We limit the Supreme Court terms included in our analysis to those that are covered by all three of these data source (1937--2015)\footnote{There were 145 cases that were listed in the SCDB but could not be matched to a case in the CourtListener data. We decided to exclude these 145 cases from our analysis}. The Supreme Court citation network from $1937 - 2015$ consists of $9,945$ cases. The breakdown of the data by the Court's Chief Justice is presented in Table \ref{tab:chiefs}. The network has a total of $111,986$ citation ties.\footnote{In order to focus on citation actions that were not intended to totally invalidate an opinion \textbf{ HERE, SOURCE FOR OVERRULED, MAJORITY ONLY?}, we exclude $315$ citations that caused the cited case to be overruled. The data on overruling citation came from \textbf{CITE}. Our results are virtually unchanged if we include the overruling citations.} The in- and outdegree distributions (i.e., the distributions of the number of citations sent and received by cases, respectively), are visualized in Figure \ref{degree_dist}. The maximum indegree (i.e., number of cases citing to a case) is $230$ and the maximum outdegree (i.e., number of cases cited by a case) is $162$. The majority of cases cite to and/or are cited by twenty or fewer other cases.

@@

\subsection{Correlates of Legal Citations}

Theorizing that legal citations arise through network interdependence processes represents a potential contribution to the literature. To demonstrate the value of our contribution we must also incorporate established non-network processes that have been found to affect legal citations. We incorporate these processes through the specification of $h_{covariate}$  terms to be included in the c-TERGM.

We incorporate a statistic to model the degree to which cases cite those that are similar in terms of the ideological positions of the justices who supported the decision. We account for this effect following \citet{spriggs2001explaining}, who find that cases are more likely to be overruled when the Court is ideologically distant from the median justice in the majority coalition that decided the case. \citet{clark2010locating} estimates a latent coordinate model of Supreme Court opinions based on the network of case-to-case citations. They find that the majority opinion falls at the ideal point of the median member of the majority coalition in the case. We include a covariate term in which $X_{ij}$ is the absolute difference between the Martin-Quinn scores \citep{martin2002dynamic} of the median justices in the majority coalitions for cases $i$ and $j$. We expect this variable to have a negative effect, which would correspond to cases citing those to which they are ideologically similar.%WE COULD USE THIS DATASET AS A SECOND APPLICATION

We include two sets of dummy variables that account for the issue areas of cases. In one set of dummy variable---sender intercepts---the variable $X^{ij}$ indicates the issue area of the sending case ($i$). In the set of receiver intercepts $X^{ij}$ indicates the issue of the receiver case. Issue area data comes from the Supreme Court Database (SCDB) \citep{spaeth2014supreme}. We include these variables because \citet{cross2010determinants} finds that the number of citations to Supreme Court opinions depends heavily on the issue area of the case. Table \ref{issue_area_coding} states the different issue areas.

@@

\begin{itemize}
\item mention new variable: overruled cases
\item mention new variable: nodemix(same issue area)
\item mention new variable:Sender/receiver effect for judge who wrote majority opinion
\item Same justice homophily with time trend
\item Argue/justify why certain covariates are included
\end{itemize}

\textit{Comment from Benjamin Kassow: In section 5, I think the discussion is mostly good, but could be organized a bit differently. 
I think the discussion of the random intercepts is a bit confusing early on (it is explained more effectively later). }
%%%% DONT DELETE THESE COMMENTS, INCLUDE IDEAS FOR ADDITIONAL COVARIATES %%%%%

>>>>

 \section{Empirical Analysis}
Our three data sources for this study include the Supreme Court Database (SCDB) \citep{spaeth2014supreme}, Martin-Quinn scores \citep{martin2002dynamic}, and Supreme Court citation data provided by the CourtListener Free Law Project \citep{CourtListener}. In the next section we explain the variables we construct using these data sources. We limit the Supreme Court terms included in our analysis to those that are covered by all three of these data source (1937--2015)\footnote{There were 145 cases that were listed in the SCDB but could not be matched to a case in the CourtListener data. We decided to exclude these 145 cases from our analysis}. The Supreme Court citation network from $1937 - 2015$ consists of $9,945$ cases. The breakdown of the data by the Court's Chief Justice is presented in Table \ref{tab:chiefs}. The network has a total of $111,986$ citation ties.\footnote{In order to focus on citation actions that were not intended to totally invalidate an opinion, we exclude $315$ citations that caused the cited case to be overruled. The data on overruling citation came from \citet{senate2016constitution}. Our results are virtually unchanged if we include the overruling citations.} The in- and outdegree distributions (i.e., the distributions of the number of citations sent and received by cases, respectively), are visualized in Figure \ref{degree_dist}. The maximum indegree (i.e., number of cases citing to a case) is $230$ and the maximum outdegree (i.e., number of cases cited by a case) is $162$. The majority of cases cite to and/or are cited by twenty or fewer other cases.

@@

\subsection{c-TERGM specification}

The c-TERGM we estimate includes two classes of terms---one that captures the effects of covariates on tie formation, and another that captures the complex dependence processes that we expect to observe in the Supreme Court Citation Network. We describe our model specification by defining the terms within these two classes.

\subsubsection{Covariate terms}

Covariate effects are accounted for in the c-TERGM via the term that is used to specify the effect of covariates in other ERGM family models, as
$$h_{covariate}(C_t,C_{<t},X) =  \sum_{ij} C_{ij,t}X_{ij}.$$ Since $C_{ij,t}$ is a binary indicator of whether case $i$ cites case $j$, $h_{covariate}$ amounts to the sum of covariate values among directed dyads for which we observed a citation. The dyadic interpretation of the coefficient attributed to this term is the change in the log odds of a tie from $i$ to $j$ given a unit increase in $X_{ij}$ (i.e., exactly the interpretation of the effect of a covariate in logistic regression). We include several exogenous covariates based on this standard statistic formulation.

The first covariate (i.e., $X$) we incorporate into the model accounts for the degree to which cases cite those that are similar in terms of the ideological positions of the justices who supported the decision. We account for this effect following \citet{spriggs2001explaining}, who find that cases are more likely to be overruled when the Court is ideologically distant from the median justice in the majority coalition that decided the case. \citet{clark2010locating} estimates a latent coordinate model of Supreme Court opinions based on the network of case-to-case citations. They find that the majority opinion falls at the ideal point of the median member of the majority coalition in the case. We include a covariate term in which $X_{ij}$ is the absolute difference between the Martin-Quinn scores \citep{martin2002dynamic} of the median justices in the majority coalitions for cases $i$ and $j$. We expect this variable to have a negative effect, which would correspond to cases citing those to which they are ideologically similar.%WE COULD USE THIS DATASET AS A SECOND APPLICATION

We include three sets of dummy variables that account for the issue areas of cases. In the first set of dummy variables---sender intercepts---the variable $X^{ij}$ indicates the issue area of the sending case ($i$) (e.g., for the Criminal Procedure indicator, $X_{ij} = 1$ if case $i$ is a criminal procedure case, and $0$ otherwise). The second set of dummy variables represent receiver issue area intercepts in which $X^{ij}$ indicates the issue of the receiver case. The third set of dummy variables account for the rate of citation within each issue area and between each combination of issue areas.  There is a separate dummy variable for each directed pair of issue areas. For example the A$\rightarrow$B dummy variable $X_{ij} = 1$ if case $i$ is issue area A and case $j$ is issue area B, and 0 otherwise. Note that, given the high dimension of these dummy variables, we omit one each of the sender and recipient indicators, and 28 of the 196 mixing rate indicators. Unfortunately, this leaves these indicators largely uninterpretable, but these variables are nonetheless important to absorb variation in citations that is attributable to the legal domains of two cases in a dyad. The Issue area data comes from the Supreme Court Database (SCDB) \citep{spaeth2014supreme}. We include these variables because \citet{cross2010determinants} finds that the number of citations to Supreme Court opinions depends heavily on the issue area of the case. Table \ref{issue_area_coding} reports the issue areas covered in our data.

We also include a set of covariates that account for justice-specific variation in citations. We include two types of terms at the justice level. First, we include an indicator variable that models the rate at which justices cite themselves. This effect is modeled with a single indicator variable in which $X_{ij} = 1$ if the majority opinions in cases $i$ and $j$ were written by the same justice. Second, we use two sets of dummy variables to account for differences across justices both in terms of their tendencies to include citations in the opinions they write and to be cited by other opinions. The justice sender dummy variables indicate the identity of the justice who authored the majority or plurality opinion of case $i$. The justice recipient dummy variables indicate the identity of the justice who authored the majority or plurality opinion of case $j$. This set of variables is motivated by the use of citations by \citet{kosma1998measuring} to measure the influence of individual Supreme Court justices.

@@

The final control variable we include in the model is one in which $X_{ij} = 1$ if case $j$ was overruled prior to the term in which case $i$ was decided. This variable, quite simply, models the effect of being overruled on the rate of citation to a case after the overruling citation. \citet{fowler2008authority} find that citations to a case drop off quickly after the case has been overruled.

\subsubsection{Dependence terms}

%\textit{Comment from Benjamin Kassow: In section 5, I think the discussion is mostly good, but could be organized a bit differently. 
%I think the discussion of the random intercepts is a bit confusing early on (it is explained more effectively later). }
%%%% DONT DELETE THESE COMMENTS, INCLUDE IDEAS FOR ADDITIONAL COVARIATES %%%%%
